Spatial Generalized Linear Models in {R} Using {spmodel}

### 2025-08-21 inst/manuscript/manuscript.tex
some updates

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  article]{jss}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{orcidlink,thumbpdf,lmodern}

\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}{}
\makeatother
\makeatletter
\makeatother
\makeatletter
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, sharp corners, interior hidden]}{\end{tcolorbox}}\fi
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Spatial Generalized Linear Models in  Using },
  pdfauthor={Michael Dumelle; Jay M. Ver Hoef; Matt Higham},
  pdfkeywords={Autoregressive Model, Geostatistics, Spatial
Covariance, Spatially-Explicit Model, Statistical Model},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

%% -- Article metainformation (author, title, ...) -----------------------------

%% Author information
\author{Michael Dumelle~\orcidlink{0000-0002-3393-5529}\\United States\\
Environmental Protection Agency \And Jay M. Ver
Hoef~\orcidlink{0000-0003-4302-6895}\\National Oceanic and\\
Atmospheric Administration \AND Matt
Higham~\orcidlink{0009-0006-4217-625X}\\St.~Lawrence University}
\Plainauthor{Michael Dumelle, Jay M. Ver Hoef, Matt
Higham} %% comma-separated

\title{Spatial Generalized Linear Models in \proglang{R} Using
\pkg{spmodel}}
\Plaintitle{Spatial Generalized Linear Models in
Using} %% without formatting

%% an abstract and keywords
\Abstract{Generalized linear models (GLMs) describe a non-normal
response variable that may be binary, count, skewed, or a proportion.
Typically, observations in a GLM are assumed independent of one another.
For spatial data, this independence assumption is impractical, as nearby
locations tend to be more similar than locations far apart. The
\pkg{spmodel} \proglang{R} package provides tools to fit GLMs that
incorporate spatial autocorrelation (i.e., spatial generalized linear
models, or SPGLMs). SPGLMs are fit in \pkg{spmodel} using a novel
application of the Laplace approximation via \fct{spglm} for
point-referenced data or \fct{spgautor} for areal (i.e., lattice), data.
\fct{spglm} and \fct{spgautor} closely resemble \fct{glm} from base
\proglang{R} but include arguments that control the spatial
autocorrelation structure. \pkg{spmodel} has many helper functions for
model inspection and diagnostics, some of which leverage other
\proglang{R} packages like \pkg{broom} and \pkg{emmeans}. \pkg{spmodel}
has tools to make predictions of the latent spatial-mean process at
unobserved locations. \pkg{spmodel} also provides many advanced features
like accommodating geometric anisotropy and nonspatial random effects,
simulating spatially autocorrelated data, and more. Here we use
\pkg{spmodel} to illustrate the modeling of binary, count, skewed and
proportion response variables from several point-referenced and areal
data sets.}

%% at least one keyword must be supplied
\Keywords{Autoregressive Model, Geostatistics, Spatial
Covariance, Spatially-Explicit Model, Statistical Model}
\Plainkeywords{Autoregressive Model, Geostatistics, Spatial
Covariance, Spatially-Explicit Model, Statistical Model}

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}
%% \setcounter{page}{1}
%% \Pages{1--xx}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
Michael Dumelle\\
Pacific Ecological Systems Division\\
200 SW 35th St\\
Corvallis OR\\
E-mail: \email{dumelle.michael@epa.gov}\\
\\~
Jay M. Ver Hoef\\
Alaska Fisheries Science Center\\
\\~
Matt Higham\\
Department of Math, Computer Science, and Statistics\\
\\~

@@

\begin{document}
\maketitle

@@

Binary, count, proportion, and skewed data are ubiquitous in practice.
These data types are naturally modeled using a generalized linear model
(GLM) framework
\citep{nelder1972generalized, mccullagh1989generalized, myers2012generalized, faraway2016extending}.
In a GLM, a response variable \(\text{y}\) belongs to a particular
statistical distribution with mean \(\mu\). For example, \(\text{y}\)
may be distributed as a Poisson random variable with some mean \(\mu\).
Based on the response distribution assumed for \(\text{y}\), GLMs link a
function of \(\mu\) to explanatory variables via a link function:
\begin{equation}\label{eq-glm}
f(\boldsymbol{\mu}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta},
\end{equation} where for a sample size \(n\), \(\boldsymbol{\mu}\) is
the mean of an \(n \times 1\) response vector \(\mathbf{y}\),
\(f(\boldsymbol{\mu})\) is the link function that connects
\(\boldsymbol{\mu}\) to \(\mathbf{w}\), \(\mathbf{X}\) is the
\(n \times p\) design matrix of explanatory variables, and
\(\boldsymbol{\beta}\) is the \(p \times 1\) vector of fixed effects.
The mean, \(\boldsymbol{\mu}\), is usually constrained in some way
(e.g., positive) that depends on the distribution assumed for
\(\mathbf{y}\), but \(\mathbf{w}\) is unconstrained. The parameters in
Equation\(~\)\eqref{eq-glm} are typically estimated via maximum
likelihood (e.g., iteratively reweighted least squares)
\citep{chambers1992S}. The \fct{glm} function is commonly used to fit
GLMs in the \proglang{R} programming language \citep{rcore2024}.

@@

The \pkg{spmodel} \proglang{R} package provides tools for fitting
spatial statistical models and making predictions at unobserved
locations \citep{dumelle2023spmodel}. A recent major update (v0.4.0) to
\pkg{spmodel} added SPGLM (Equation\(~\)\ref{eq-spglm}) support for
binary, count, skewed, and proportion response variables
(Table\(~\)\ref{table-links}), greatly expanding the class of models
that \pkg{spmodel} makes accessible to practitioners.

\begin{table}

\centering{

\centering
\begin{tabular}{||c|ccc||}
 \hline
 Family & Link Function & Link Name & Data Type  \\ [0.5ex]
 \hline \hline
 Binomial & $f(\mu) = \log(\mu / (1 - \mu))$ & Logit & Binary; Binary Count \\
 Poisson & $f(\mu) = \log(\mu)$ & Log & Count \\
 Negative Binomial & $f(\mu) = \log(\mu)$ & Log & Count \\
 Beta & $f(\mu) = \log(\mu / (1 - \mu))$ & Logit & Proportion \\
 Gamma & $f(\mu) = \log(\mu)$ & Log & Skewed \\
 Inverse Gaussian & $f(\mu) = \log(\mu)$ & Log & Skewed \\
 \hline\hline
\end{tabular}

\caption{\label{table-links}SPGLM response distributions and their link functions and data types.}

\end{table}%

SPGLMs for point-referenced data are called geostatistical GLMs and are
fit using \fct{spglm}. Data are point-referenced when the elements in
\(\mathbf{y}\) are observed at point-locations indexed by x-coordinates
and y-coordinates on a spatially continuous surface with an infinite
number of locations (e.g., point locations in a field). SPGLMs for areal
data are called spatial autoregresive models and are fit using
\fct{spgautor}. Data are areal (i.e., lattice) when they are part of a
finite network of polygons whose connections are indexed by a
neighborhood structure (e.g., states in a country).

Several other \proglang{R} packages exist for analyzing SPGLMs. The
\pkg{brms} \citep{burkner2017brms}, \pkg{carBayes}
\citep{lee2013carbayes}, \pkg{ngspatial} \citep{hughes2020ngspatial},
\pkg{R-INLA} \citep{lindgren2015bayesian}, \pkg{spBayes}
\citep{finley2007spbayes}, and \pkg{spNNGP} \citep{finley2002spnngp}
packages take a Bayesian approach, either directly sampling from
posterior distributions of parameters (e.g., using MCMC) or
approximating them. A benefit of Bayesian approaches is that prior
information can be incorporated and uncertainty quantification of
parameter estimates is straightforward. However, Bayesian approaches,
especially those using MCMC, tend to be computationally expensive. In
order to reduce computation time, many of these packages work with the
precision matrix instead of the covariance matrix so that inverting
matrices (a computational bottleneck) is not required. For example,
\pkg{R-INLA} uses the precision matrix (often directly modeled as with
autoregressive models) and tends to be very fast. Working with precision
matrices, however, can be more restrictive and less intuitive than
working directly with the covariance matrix. The \pkg{FRK}
\citep{sainsbury2024modeling}, \pkg{glmmTMB} \citep{brooks2017glmmtmb},
\pkg{hglm} \citep{ronnegard2010hglm}, \pkg{mgcv}
\citep{wood2017generalized}, and \pkg{spaMM} \citep{rousset2014spamm}
packages directly use Laplace, quasi-likelihood, or reduced-rank
approaches to estimate parameters. These direct approaches tend to be
computationally efficient, as they don't rely on MCMC sampling. In
contrast to the Bayesian approach, a drawback of these direct approaches
is that prior information cannot be formally incorporated and covariance
parameter uncertainty is challenging to quantify. SPGLMs in
\pkg{spmodel} are most similar to those in \pkg{glmmTMB} --
\pkg{spmodel} uses analytical solutions for maximizing the likelihood
(which we describe in the next section), while \pkg{glmmTMB} uses
automatic differentiation for maximizing the likelihood (which tends to
be much slower than analytical solutions).

Missing from the aforementioned \proglang{R} packages is the complete
set of tools for SPGLMs that \pkg{spmodel} provides. Importantly, the
\fct{spglm} and \fct{spautor} functions act as a spatial analogue to the
familiar \fct{glm} from base \proglang{R}, making the transition from
GLMs to SPGLMs relatively seamless. \pkg{spmodel} leverages many
commonly used \proglang{R} generics like \fct{summary} to better
understand fitted models. Six GLM families (Table\(~\)\ref{table-links})
and 20 spatial covariance (or precision) functions are supported. Also
available are functions for data visualization, model fitting, model
summaries, model diagnostics, model comparison, and prediction, all
crucial components of a data analysis. Through extra function arguments,
\pkg{spmodel} supports advanced features like geometric anisotropy,
non-spatial random effects, methods for large data sets, and more.
Importantly, \pkg{spmodel} extends popular packages like \pkg{broom}
\citep{robinson2021broom, kuhn2022tidy} and \pkg{emmeans}
\citep{lenth2024emmeans}. \pkg{spmodel} also provides function for
simulating SPGLM data (e.g., \fct{sprbinom}, \fct{sprpois}).

The rest of this article is organized as follows. In
Section~\ref{sec-spglm}, we provide some background for the SPGLM
fitting and prediction routines in \pkg{spmodel}. In
Section~\ref{sec-applications}, we provide several applications of
\pkg{spmodel} to spatial binary, count, skewed and proportion data with
both point-referenced and areal supports. And in
Section~\ref{sec-discussion}, we end with a discussion synthesizing
\pkg{spmodel}'s contributions to the analysis of SPGLMs in \proglang{R}.

@@

GLMs with random effects are often written hierarchically
\citep{lee1996hierarchical, bolker2009generalized, wood2017generalized}.
\pkg{spmodel} leverages this hierarchical structure and uses a novel
application of the Laplace approximation \citep{ver2024marginal} to fit
models. This marginal approach is quite flexible, accommodating a wide
range of possible dependence structures and formally maximizing a
likelihood. Maximizing a likelihood yields convenient likelihood-based
statistics like AIC \citep{akaike1974new}, AICc
\citep{hoeting2006model}, BIC \citep{schwarz1978estimating}, deviance
\citep{mccullagh1989generalized}, and likelihood ratio tests for model
comparison, a benefit compared to quasi-likelihood
\citep{wedderburn1974quasi, breslow1993approximate} or pseudo-likelihood
approaches \citep{wolfinger1993generalized}, which only specify the
first two moments of a distribution. \citet{ver2024marginal} provide
thorough context for the marginal approach and its associated details,
but next we provide an short overview of the methodology.

@@

The \fct{spglm} (for point-referenced data) and \fct{spgautor} (for
areal data) functions in \pkg{spmodel} fit SPGLMs using the Laplace
approximation outlined in Section~\ref{sec-spglm}. Both \fct{spglm} and
\fct{spgautor} generally require the following four arguments:
\texttt{formula}, a formula that describes the relationship between the
response variable and explanatory variables; \texttt{family}, the
response distribution (which can be \texttt{binomial}, \texttt{poisson},
\texttt{nbinomial}, \texttt{Gamma}, \texttt{inverse.gaussian}, or
\texttt{beta}); \texttt{data}, the data frame that holds the variables
in \texttt{formula} as well as spatial locations; and
\texttt{spcov\_type}, the spatial covariance type. The first three
arguments are shared by \fct{glm}; thus, the transition from GLMs to
SPGLMs requires only one additional argument: \texttt{spcov\_type}. The
\fct{spglm} spatial covariance types measure dependence as a function of
Euclidean distance among observations; an example is the exponential
spatial covariance: \begin{equation}\label{eq-spcov-exp}
  \boldsymbol{\Sigma} = \sigma^2_{de} \exp(-\mathbf{H}/\phi) + \sigma^2_{ie}\mathbf{I},
\end{equation} where \(\mathbf{H}\) is a matrix of pairwise distances
among all observations. \fct{spglm} currently supports 18 distinct
spatial covariance functions.

The \fct{spgautor} spatial covariance types measure dependence as a
function of neighborhood distance among observations; an example is the
simultaneous autoregressive covariance matrix:
\begin{equation}\label{eq-spcov-sar}
  \boldsymbol{\Sigma} = \sigma^2_{de} [(\mathbf{I} - \phi \mathbf{W})(\mathbf{I} - \phi \mathbf{W})^T]^{-1} + \sigma^2_{ie}\mathbf{I},
\end{equation} where \(\mathbf{W}\) is a matrix that represents the
neighborhood structure among all observations. \fct{spgautor} currently
supports two distinct spatial covariance functions.

In the rest of this section, we use \pkg{spmodel} to study binary,
count, skewed, and proportion response variables that are either
point-referenced or areal. We use \pkg{spmodel} for all parts of the
data analysis, from estimation to inference to model diagnostics to
prediction. We first describe core \pkg{spmodel} functionality in an
application to binary data, while additional analyses highlight count,
skewed, and proportion data as well as some additional \pkg{spmodel}
features. Before proceeding, load \pkg{spmodel} into the current
\proglang{R} session:

\begin{verbatim}
R> library("spmodel")
\end{verbatim}

@@

The \texttt{moose} data in \pkg{spmodel} contain information on moose
presence in Alaska. They are an \texttt{sf} object, a special data frame
that is supplemented with spatial information using the \pkg{sf} package
\citep{pebesma2018sf}. The first few rows look like:

\begin{verbatim}
R> head(moose)
\end{verbatim}

\begin{verbatim}
Simple feature collection with 6 features and 4 fields
Geometry type: POINT
Dimension:     XY
Bounding box:  xmin: 281896.4 ymin: 1518398 xmax: 311325.3 ymax: 1541016
Projected CRS: NAD83 / Alaska Albers
# A tibble: 6 x 5
   elev strat count presence           geometry
  <dbl> <chr> <dbl> <fct>           <POINT [m]>
1  469. L         0 0        (293542.6 1541016)
2  362. L         0 0        (298313.1 1533972)
3  173. M         0 0        (281896.4 1532516)
4  280. L         0 0        (298651.3 1530264)
5  620. L         0 0        (311325.3 1527705)
6  164. M         0 0        (291421.5 1518398)
\end{verbatim}

>>>>

\documentclass[
]{jss}

%% recommended packages
\usepackage{orcidlink,thumbpdf,lmodern}

\usepackage[utf8]{inputenc}

\author{
Michael Dumelle~\orcidlink{0000-0002-3393-5529}\\United States\\
Environmental Protection Agency \And Jay M. Ver
Hoef~\orcidlink{0000-0003-4302-6895}\\Alaska Fisheries\\
Science Center \And Matt
Higham~\orcidlink{0009-0006-4217-625X}\\St.~Lawrence University
}
\title{Spatial Generalized Linear Models in {R} Using {spmodel}}

\Plainauthor{Michael Dumelle, Jay M. Ver Hoef, Matt Higham}
\Plaintitle{Spatial Generalized Linear Models in R Using spmodel}
\Shorttitle{SPGLMs in {R} Using {spmodel}}

\Abstract{
Generalized linear models (GLMs) describe a non-normal response variable
that may be binary, count, skewed, or a proportion. Typically,
observations in a GLM are assumed independent of one another. For
spatial data, this independence assumption is impractical, as nearby
locations tend to be more similar than locations far apart. The
{spmodel} {R} package provides tools to fit GLMs that incorporate
spatial autocorrelation (i.e., spatial generalized linear models, or
SPGLMs). SPGLMs are fit in {spmodel} using a novel application of the
Laplace approximation via {spglm} for point-referenced data or
{spgautor} for areal (i.e., lattice), data. {spglm} and {spgautor}
closely resemble {glm} from base {R} but include arguments that control
the spatial autocorrelation structure. {spmodel} has many helper
functions for model inspection and diagnostics, some of which leverage
other {R} packages like {broom} and {emmeans}. {spmodel} has tools to
make predictions of the latent spatial-mean process at unobserved
locations. {spmodel} also provides many advanced features like
accommodating geometric anisotropy and nonspatial random effects,
simulating spatially autocorrelated data, and more. Here we use
{spmodel} to illustrate the modeling of binary, count, skewed and
proportion response variables from several point-referenced and areal
data sets.
}

\Keywords{autoregressive model, geostatistical model, spatial
autocovariance, spatial autocorrelation}
\Plainkeywords{autoregressive model, geostatistical model, spatial
autocovariance, spatial autocorrelation}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
    Michael Dumelle\\
    United States\\
Environmental Protection Agency\\
    200 SW 35th St\\
Corvallis, OR, 97330\\
  E-mail: \email{Dumelle.Michael@epa.gov}\\

@@

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{amsmath,amsfonts,amssymb}
\usepackage{bm, bbm}
\usepackage{lineno}
\usepackage{caption, subcaption}

\begin{document}

@@

Binary, count, proportion, and skewed data are ubiquitous in practice.
These data types are naturally modeled using a generalized linear model
(GLM) framework
\citep{nelder1972generalized, mccullagh1989generalized, myers2012generalized, faraway2016extending}.
In a GLM, a response variable \(\text{y}\) belongs to a particular
statistical distribution with mean \(\mu\). For example, \(\text{y}\)
may be distributed as a Poisson random variable with some mean \(\mu\).
Based on the response distribution assumed for \(\text{y}\), GLMs link a
function of \(\mu\) to explanatory variables via a link function:
\begin{equation}\label{eq-glm}
f(\boldsymbol{\mu}) \equiv \mathbf{w} = \mathbf{X} \boldsymbol{\beta},
\end{equation} where for a sample size \(n\), \(\boldsymbol{\mu}\) is
the mean of an \(n \times 1\) response vector \(\mathbf{y}\),
\(f(\boldsymbol{\mu})\) is the link function that connects
\(\boldsymbol{\mu}\) to \(\mathbf{w}\), \(\mathbf{X}\) is the
\(n \times p\) design matrix of explanatory variables, and
\(\boldsymbol{\beta}\) is the \(p \times 1\) vector of fixed effects.
The mean, \(\boldsymbol{\mu}\), is usually constrained in some way
(e.g., positive) that depends on the distribution assumed for
\(\mathbf{y}\), but \(\mathbf{w}\) is unconstrained. The parameters in
Equation\(~\)\eqref{eq-glm} are typically estimated via maximum
likelihood (e.g., iteratively reweighted least squares)
\citep{chambers1992S}. The {glm} function is commonly used to fit GLMs
in the {R} programming language \citep{rcore2024}.

@@

The {spmodel} {R} package provides tools for fitting spatial statistical
models and making predictions at unobserved locations
\citep{dumelle2023spmodel}. A recent major update (v0.4.0) to {spmodel}
added SPGLM (Equation\(~\)\ref{eq-spglm}) support for binary, count,
skewed, and proportion response variables (Table\(~\)\ref{table-links}),
greatly expanding the class of models that {spmodel} makes accessible to
practitioners.

\begin{table}
\centering
\begin{tabular}{||c|ccc||} 
 \hline
 Family & Link Function & Link Name & Data Type  \\ [0.5ex] 
 \hline \hline
 Binomial & $f(\mu) = \log(\mu / (1 - \mu))$ & Logit & Binary; Binary Count \\
 Poisson & $f(\mu) = \log(\mu)$ & Log & Count \\
 Negative Binomial & $f(\mu) = \log(\mu)$ & Log & Count \\
 Beta & $f(\mu) = \log(\mu / (1 - \mu))$ & Logit & Proportion \\
 Gamma & $f(\mu) = \log(\mu)$ & Log & Skewed \\
 Inverse Gaussian & $f(\mu) = \log(\mu)$ & Log & Skewed \\ 
 \hline\hline
\end{tabular}
\caption{SPGLM response distributions and their link functions and data types.}
\label{table-links}
\end{table}

SPGLMs for point-referenced data are called geostatistical GLMs and are
fit using {spglm}. Data are point-referenced when the elements in
\(\mathbf{y}\) are observed at point-locations indexed by x-coordinates
and y-coordinates on a spatially continuous surface with an infinite
number of locations (e.g., point locations in a field). SPGLMs for areal
data are called spatial autoregresive models and are fit using
{spgautor}. Data are areal (i.e., lattice) when they are part of a
finite network of polygons whose connections are indexed by a
neighborhood structure (e.g., states in a country).

Several other {R} packages exist for analyzing SPGLMs. The {brms}
\citep{burkner2017brms}, {carBayes} \citep{lee2013carbayes}, {ngspatial}
\citep{hughes2020ngspatial}, {R-INLA} \citep{lindgren2015bayesian},
{spBayes} \citep{finley2007spbayes}, and {spNNGP}
\citep{finley2002spnngp} packages take a Bayesian approach, either
directly sampling from posterior distributions of parameters (e.g.,
using MCMC) or approximating them. A benefit of Bayesian approaches is
that prior information can be incorporated and uncertainty
quantification of parameter estimates is straightforward. However,
Bayesian approaches, especially those using MCMC, tend to be
computationally expensive. In order to reduce computation time, many of
these packages work with the precision matrix instead of the covariance
matrix so that inverting matrices (a computational bottleneck) is not
required. For example, {R-INLA} uses the precision matrix (often
directly modeled as with autoregressive models) and tends to be very
fast. Working with precision matrices, however, can be more restrictive
and less intuitive than working directly with the covariance matrix. The
{FRK} \citep{sainsbury2024modeling}, {glmmTMB}
\citep{brooks2017glmmtmb}, {hglm} \citep{ronnegard2010hglm}, {mgcv}
\citep{wood2017generalized}, and {spaMM} \citep{rousset2014spamm}
packages directly use Laplace, quasi-likelihood, or reduced-rank
approaches to estimate parameters. These direct approaches tend to be
computationally efficient, as they don't rely on MCMC sampling. In
contrast to the Bayesian approach, a drawback of these direct approaches
is that prior information cannot be formally incorporated and covariance
parameter uncertainty is challenging to quantify. SPGLMs in {spmodel}
are most similar to those in {glmmTMB} -- {spmodel} uses analytical
solutions for maximizing the likelihood (which we describe in the next
section), while {glmmTMB} uses automatic differentiation for maximizing
the likelihood (which tends to be much slower than analytical
solutions).

Missing from the aforementioned {R} packages is the complete set of
tools for SPGLMs that {spmodel} provides. Importantly, the {spglm} and
{spautor} functions act as a spatial analogue to the familiar {glm} from
base {R}, making the transition from GLMs to SPGLMs relatively seamless.
{spmodel} leverages many commonly used {R} generics like {summary} to
better understand fitted models. Six GLM families
(Table\(~\)\ref{table-links}) and 20 spatial covariance (or precision)
functions are supported. Also available are functions for data
visualization, model fitting, model summaries, model diagnostics, model
comparison, and prediction, all crucial components of a data analysis.
Through extra function arguments, {spmodel} supports advanced features
like geometric anisotropy, non-spatial random effects, methods for large
data sets, and more. Importantly, {spmodel} extends popular packages
like {broom} \citep{robinson2021broom, kuhn2022tidy} and {emmeans}
\citep{lenth2024emmeans}. {spmodel} also provides function for
simulating SPGLM data (e.g., {sprbinom}, {sprpois}).

The rest of this article is organized as follows. In
Section\(~\)\ref{sec-spglm}, we provide some background for the SPGLM
fitting and prediction routines in {spmodel}. In
Section\(~\)\ref{sec-applications}, we provide several applications of
{spmodel} to spatial binary, count, skewed and proportion data with both
point-referenced and areal supports. And in
Section\(~\)\ref{sec-discussion}, we end with a discussion synthesizing
\pkg{spmodel}'s contributions to the analysis of SPGLMs in {R}.

@@

GLMs with random effects are often written hierarchically
\citep{lee1996hierarchical, bolker2009generalized, wood2017generalized}.
{spmodel} leverages this hierarchical structure and uses a novel
application of the Laplace approximation \citep{ver2024marginal} to fit
models. This marginal approach is quite flexible, accommodating a wide
range of possible dependence structures and formally maximizing a
likelihood. Maximizing a likelihood yields convenient likelihood-based
statistics like AIC \citep{akaike1974new}, AICc
\citep{hoeting2006model}, BIC \citep{schwarz1978estimating}, deviance
\citep{mccullagh1989generalized}, and likelihood ratio tests for model
comparison, a benefit compared to quasi-likelihood
\citep{wedderburn1974quasi, breslow1993approximate} or pseudo-likelihood
approaches \citep{wolfinger1993generalized}, which only specify the
first two moments of a distribution. \citet{ver2024marginal} provide
thorough context for the marginal approach and its associated details,
but next we provide an short overview of the methodology.

@@

The {spglm} (for point-referenced data) and {spgautor} (for areal data)
functions in {spmodel} fit SPGLMs using the Laplace approximation
outlined in Section\(~\)\ref{sec-spglm}. Both {spglm} and {spgautor}
generally require the following four arguments: \texttt{formula}, a
formula that describes the relationship between the response variable
and explanatory variables; \texttt{family}, the response distribution
(which can be \texttt{binomial}, \texttt{poisson}, \texttt{nbinomial},
\texttt{Gamma}, \texttt{inverse.gaussian}, or \texttt{beta});
\texttt{data}, the data frame that holds the variables in
\texttt{formula} as well as spatial locations; and \texttt{spcov\_type},
the spatial covariance type. The first three arguments are shared by
{glm}; thus, the transition from GLMs to SPGLMs requires only one
additional argument: \texttt{spcov\_type}. The {spglm} spatial
covariance types measure dependence as a function of Euclidean distance
among observations; an example is the exponential spatial covariance:
\begin{equation}\label{eq-spcov-exp}
  \boldsymbol{\Sigma} = \sigma^2_{de} \exp(-\mathbf{H}/\phi) + \sigma^2_{ie}\mathbf{I},
\end{equation} where \(\mathbf{H}\) is a matrix of pairwise distances
among all observations. {spglm} currently supports 18 distinct spatial
covariance functions.

The {spgautor} spatial covariance types measure dependence as a function
of neighborhood distance among observations; an example is the
simultaneous autoregressive covariance matrix:
\begin{equation}\label{eq-spcov-sar}
  \boldsymbol{\Sigma} = \sigma^2_{de} [(\mathbf{I} - \phi \mathbf{W})(\mathbf{I} - \phi \mathbf{W})^T]^{-1} + \sigma^2_{ie}\mathbf{I},
\end{equation} where \(\mathbf{W}\) is a matrix that represents the
neighborhood structure among all observations. {spgautor} currently
supports two distinct spatial covariance functions.

In the rest of this section, we use {spmodel} to study binary, count,
skewed, and proportion response variables that are either
point-referenced or areal. We use {spmodel} for all parts of the data
analysis, from estimation to inference to model diagnostics to
prediction. We first describe core {spmodel} functionality in an
application to binary data, while additional analyses highlight count,
skewed, and proportion data as well as some additional {spmodel}
features. Before proceeding, load {spmodel} into the current {R}
session:

\begin{CodeChunk}
\begin{CodeInput}
R> library("spmodel")
\end{CodeInput}
\end{CodeChunk}

@@

The \texttt{moose} data in {spmodel} contain information on moose
presence in Alaska. They are an \texttt{sf} object, a special data frame
that is supplemented with spatial information using the {sf} package
\citep{pebesma2018sf}. The first few rows look like:

\begin{CodeChunk}
\begin{CodeInput}
R> head(moose)
\end{CodeInput}
\begin{CodeOutput}
Simple feature collection with 6 features and 4 fields
Geometry type: POINT
Dimension:     XY
Bounding box:  xmin: 281896.4 ymin: 1518398 xmax: 311325.3 ymax: 1541016
Projected CRS: NAD83 / Alaska Albers
# A tibble: 6 x 5
   elev strat count presence           geometry
  <dbl> <chr> <dbl> <fct>           <POINT [m]>
1  469. L         0 0        (293542.6 1541016)
2  362. L         0 0        (298313.1 1533972)
3  173. M         0 0        (281896.4 1532516)
4  280. L         0 0        (298651.3 1530264)
5  620. L         0 0        (311325.3 1527705)
6  164. M         0 0        (291421.5 1518398)
\end{CodeOutput}
\end{CodeChunk}

There are five columns: \texttt{elev}, the numeric site elevation
(meters); \texttt{strat} a stratification variable for sampling with two
levels, \texttt{"L"} and \texttt{"M"}, which are categorized by
landscape metrics at each site; \texttt{count}, the number of moose at
each site; \texttt{presence}, a factor that indicates whether at least
one moose was observed at each site (\texttt{0} implies no moose;
\texttt{1} implies at least one moose); and \texttt{geometry}, the NAD83
projected coordinate of each site. The \texttt{moose\_preds} data in
{spmodel} contain spatial locations at which predictions of moose
presence are desired. They are also an \texttt{sf} object with the same
projection and measurements for \texttt{elev} and \texttt{strat}.
Figure\(~\)\ref{fig-moose-data} shows the \texttt{presence} variable in
\texttt{moose} as well as the spatial locations of both \texttt{moose}
and \texttt{moose\_preds}. Moose are most common in the southwestern and
eastern parts of the domain and least common in the northwest.
